The value of early warning systems for pandemic response

### 2024-02-26 reproduce-results.qmd
sensitivity analysis results with epi and societal inputs are sensible

---
title: "results"
format: html
author: "Pedro Nascimento de Lima"
---

@@

This quarto document reproduces the results section in Nascimento de Lima et al. (2024). *The value of environmental surveillance for pandemic response*. For details on how to install dependencies, and reproduce these results, see the `README.md` file.

# Run Results

First, we instantiate the model and run it over the experimental designs used in the paper.

```{r}

@@

## Value of EWS under base-case results
{Describe dynamics and results uder base-case assumptions}

D. EWS provide meaningful value under base-case assumptions

## Conditions for high-value early warning 
{Describe conditions under which value is lower or higher}
D. Value of EWS hinges on timeliness of and adherence to NPIs. 
There was a

## EWS and policy coordination
D. EWS alleviate lack of policy coordination. 
There is a concern that EWS data will need to be integrated to be useful. However, we find that policy coordination and EWS act as substitutes. Policy coordination h

## Marginal Value of EWS
D. EWS deployment also face decreasing marginal returns
The value of EWS decreases as coverage increases.

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

>>>>

---
title: "The value of early warning systems for pandemic response"
format:
  html:
    self-contained: true
execute: 
  echo: true
  warning: false
  error: false
author: "Pedro Nascimento de Lima"
---

@@

This `quarto` document can be used to reproduce the results presented in *Nascimento de Lima et al. (2024). The value of environmental surveillance for pandemic response*. For details on how to install dependencies, see the `README.md` file. Before attempting to run this document, install all dependencies and run the `run_analysis.R` script. You should be able to reproduce all figures and tables from our paper using this document on a modern laptop within 30 minutes.

# Running the model

The following code block uses results stored in the `./output/r.rds` file or runs the script that generates this file.

```{r}
# The r object contains the results objects used in this file
if(!file.exists("./output/r.rds")) {
  source("./R/scripts/run_analysis.R")
} else {
  source("./R/library.R")
  r <- readRDS("./output/r.rds")
}

@@

## Value of EWS under baseline assumptions

<!--EWS systems provide meaningful value under base-case assumptions representing a hypothetical new pandemic with parameters mirroring those of SARS-Cov-2. -->
Assuming no NPIs interventions are put in place (No NPIs scenario), a new COVID-19-like pandemic would produce significant damage, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` deaths per 100,000 people [mean (95% prediction interval of stochastic replications)] over a one-year time-frame (Table 1). The unmitigated pandemic would infect `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "epi_size")` percent of the population, causing `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_illness")` dollars per person due to ilness, and `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH_deaths")` dollars due to deaths. The total health cost would be projected at `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "CH")` dollars per person.

If policymakers follow the base-case NPI policy simulated in this paper, the death toll of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` deaths per 100,000 population if no enhanced early-warning system (EWS) is adopted (NPIs w/o EWS scenario). As it was the case during the COVID-19 pandemic, such an achievement would involve a substantial public health effort. Under baseline assumptions, this effort would require `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` days of public health interventions, of which `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` days are spent under the most-restrictive policy intervention level. We project this effort to cost on the order of `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` per person. Based on those assumptions, the total cost of the pandemic would be reduced from `r pull_estimate(r$table_1_long, Scenario == "No NPIs", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` per person.

```{r}
#| label: tbl-1
#| tbl-cap: Base-case results
gt::gt(r$table_1)
```

Notes: All costs are expressed as 2020 dollars per person. All outcomes were computed over a one-year time-frame.

Under base-case assumptions, EWS provide net-positive value to society because they enhance the effectiveness of NPIs (i.e., preventing illness and death) more than they increase the burden of additional interventions introduced due to the early warning. Even an early warning system that provides only a 2-day early warning relative to a baseline pandemic response system (NPI + 2-day EWS scenario) is projected to reduce the pandemic death toll from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "deaths_per_100k")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 2-day EWS", variable == "deaths_per_100k")` deaths per 100,000 population. Health benefits increase if the system can provide earlier  warning. A system that provides a 5-day early warning (NPIs + 5-day EWS scenario) would be projected to result in `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "deaths_per_100k")` deaths per 100,000 population, and a 10-day early warning would result in `r pull_estimate(r$table_1_long, Scenario == "NPIs + 10-day EWS", variable == "deaths_per_100k")` deaths per 100,000 people.

The use of early warning would however result in earlier public health intervention, increasing the number of days of under NPIs from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L1plus_days")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "L1plus_days")` days for a system providing a 5-day early warning signal. That said, the number of days at the *maximum* intervention level (i.e., a lockdown) was projected to *decrease* with the use of the early warning system, as earlier action prevented surges from reaching alarming levels that prompted blunt interventions. For example, a system that provided a 5-day early warning reduced the number of days under the maximum intervention level from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "L5_days")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "L5_days")` days. Still, the net effect of early warning system was to marginally increase NPI intervention costs, from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "CNPI")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "CNPI")` dollars per person, assuming a 5-day early warning. Health and economic outcomes weighted, the early warning system reduced the total pandemic costs from `r pull_estimate(r$table_1_long, Scenario == "NPIs w/o EWS", variable == "C")` to `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "C")` dollars per person. The difference between the two costs can be interpreted as the net-monetary benefit of the system, which is estimated at `r pull_estimate(r$table_1_long, Scenario == "NPIs + 5-day EWS", variable == "NMB")` dollars per person in the first year of the pandemic.

<!--
Points from this section for discussion:
* **EWS provide meaningful value under base-case assumptions.** Net-monetary benefit from surveillance system is on the order of 1,000 dollars per person over a one-year time-frame. This outweighs any available estimate of costs.
* **Base-case results align with observed COVID-19 data.** US obseved on the order of 160 deaths per 100,000 people during the first year of the pandemic, and many states.
-->

## Value of EWS system depends on epidemiological and societal parameters

### Effect of epidemiological parameters

A pathogen that is 50% more transmissible than wild-type SARS-Cov-2 would pose a greater challenge for pandemic response, causing `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 people even if mitigated with the same NPI policy considered in our base-case scenario (Supplementary Table 2). Such a pandemic would cause a substantial societal cost of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "C", NMB_comparator == T)` dollars per person. In those conditions, the EWS would be particularly valuable, reducing the death toll of the pandemic to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == F)` per 100,000 people, and would be projected to provide a NMB of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. Conversely, a pathogen that is 50% less transmissible than SARS-Cov-2 would result in a lower death toll of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` without the EWS system, reducing the net-monetary benefit of the system to `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person.

The value of the EWS system also depends on the infection fatality rate of the pathogen, but the net-benefit of the EWS would be positive even for a pathogen that is half as deadly as SARS-Cov-2.

```{r}
#| label: fig-1
#| fig-cap: Base-case results
#| column: screen
#| fig-format: svg
#| out-width: 100%
plot(r$fig_A_epi)
```

## Effect of economic parameters

## Conditions for high-value pandemic response EW systems

{Describe conditions under which value is lower or higher}

Points from this section for discussion:

* Value of EWS hinges on:
  + timeliness of decision-making
  + adherence to NPIs
  + relative costs/severity of disease

# If we have time to look at those other cases:

## EWS and policy coordination
EWS alleviate lack of policy coordination. 
There is a concern that EWS data will need to be integrated to be useful. However, we find that policy coordination and EWS act as substitutes.

## If we have time: Marginal Value of EWS
EWS deployment also face decreasing marginal returns. The value of EWS decreases as coverage increases.
